Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-18 paper/main.tex
update paper

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such
as Mallows~\cite{FliVer1986}, that have already been been used in optimization
problems successfully, with the most prominent examples being estimation of
distribution algorithms
(EDAs)~\cite{CebIruMen2014eda,CebMenLoz2015mallows}. ACO may also be considered
a type of EDA, since it builds a probability distribution model from which
solutions are sampled with a bias towards the best solutions evaluated so
far. These classical EDAs  assume that function
evaluations are cheap and/or the fitness function is white-box.\MANUEL{Do EDAs really assume that the fitness function is white-box? ACO doesn't: They are hybridized with local search }
%The proposed UMM drastically differs from the above algorithms.

In contrast, UMM aims at learning as much as possible with a limited number of
function evaluations. Moreover, unlike EDAs, UMM does not rely on a fixed-size
population of solutions that evolves over ``generations''. Instead, taking
inspiration from Bayesian optimizers, UMM considers a sample of permutations
during the whole execution that is increased by one new permutation at each
iteration. The sample is used to learn a Mallows model from which a single new
permutation is sampled. The model is learned by using
uBorda~\cite{IruLobPer2020arxiv}, a generalization of the well-known rank
aggregation algorithm Borda~\cite{Borda1781}. Rank aggregation algorithms
summarize in a single permutation the information of a sample of permutations,
and they can be seen as functions to compute the \emph{sample mean}. The uBorda
algorithm computes a \textit{weighted sample mean} and has been shown to return
the correct estimator with high probability on streaming
contexts~\cite{IruLobPer2020arxiv}.
%
Finally, the balance between intensification and exploration is dynamically
adapted by controlling the variance of the probabilistic Mallows model.

@@

\subsection{Distributions over permutations}
Distributions over permutations~\cite{CriFliVer1996rankings} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}\Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}. The algorithm also implements the following strategy to avoid sampling the same points more than once, which will be more likely at the explotation phase. The strategy is as follows. Instead of sampling just one permutation from the MM, the algorithm generates $budgetMM$ permutations, $\Sigma$. Then it selects  for evaluation the farest to the sample, i.e.,  $\sigma\in \Sigma$ such that $\sigma = \arg\max\arg\min d(\sigma, \pi)$ for every $\pi\in S$.

@@

@@

\subsubsection{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO
and initially samples $\minit=10$ solutions uniformly. The other two parameters
of UMM are $r_1$ and $r_2$, which determine the balance between exploration and
exploitation. In the experimental section we will study their impact. implemented UMM using Python 3.

@@

\subsubsection{Datasets}
%
For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbf{I}\{\sigma(i)<\sigma(j)\}$, where $\mathbf{I}\colon S \times S \to \{0,1\}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}

@@

In the following paragraphs we analyze the result over real-world instances. %
Figure~\ref{fig:loplib} shows the results on the instances taken from LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
The type of instances from the synthetic LOP instances, and, thus, so does the
landscape. However, UMM shows a very similar behavior: there is a (close to)
linear decrease of the fitness of new solutions over each new evaluation. This
suggests that UMM is able to model the fitness landscape. Regarding CEGO, we
observe a similar behavior for the first 50-100 evaluations. At this point,
CEGO reaches the time limit of 3 days, so no observations can be made for
longer budgets.  Overall, the improvement from the first (random) to the last
solutions is 2-3 times larger in UMM than CEGO. %
\MANUEL{But this is a bit misleading because UMM runs for 4 times more
  evaluations, perhaps we should also measure up to the evaluation that CEGO
  reaches so we report both values?}

@@

\MANUEL{UMM is able to model the LOP landscape
  at least as well as CEGO as shown by the plots where UMM shows a linear
  decrease in fitness.}  
  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
  luego que addemás es más rápido} 
  \MANUEL{Pero UMM es peor en rec13 y nug12, y
  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
  entonces la diferencia en velocidad no importaría. } 
The time required to train the surrogate-model of \ekhine{esta establecida la complexity de CEGO???}{UMM increases linearly with the permutation length $n$, while in }
CEGO  it sharply
increases with $n$, requiring 
sometimes more than hour per fitness evaluation as shown in the experimental section. 
Hence, UMM is a computationally
feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here show that, despite the intrinsic difficulty of the
expensive black-box combinatorial scenario, there are still significant
advances yet to be made. Thus, we hope that this paper will motivate further
research. % on such problems.
\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}
A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's-$\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

@@

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
  \paragraph*{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
  \vspace*{-1em}
\paragraph*{Acknowledgements.}
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
% MANUEL: We should also thank the COST ACTION: Ask Carlos?
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable} and to Thomas Stützle (Université libre de Bruxelles) for suggesting works
discussing differences between QAPLIB instances.
\end{smaller}

>>>>

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such
as Mallows~\cite{FliVer1986}, that have already been been used in optimization
problems successfully, with the most prominent examples being estimation of
distribution algorithms
(EDAs)~\cite{CebIruMen2014eda}. ACO may also be considered
a type of EDA, since it builds a probability distribution model from which
solutions are sampled with a bias towards the best solutions evaluated so
far. These classical EDAs  assume that function
evaluations are cheap and/or the fitness function is white-box.%\MANUEL{Do EDAs really assume that the fitness function is white-box? ACO doesn't: They are hybridized with local search }
%The proposed UMM drastically differs from the above algorithms.

In contrast, UMM aims at learning as much as possible with a limited number of
function evaluations. Moreover, unlike EDAs, UMM does not rely on a fixed-size
population of solutions that evolves over ``generations''. Instead, taking
inspiration from Bayesian optimizers, UMM considers a sample of permutations
during the whole execution that is increased by one new permutation at each
iteration. The sample is used to learn a Mallows model from which a single new
permutation is sampled. The model is learned by using
uBorda~\cite{IruLobPer2020arxiv}, a generalization of the well-known rank
aggregation algorithm Borda~\cite{AliMei2011kemeny}. Rank aggregation algorithms
summarize in a single permutation the information of a sample of permutations,
and they can be seen as functions to compute the \emph{sample mean}. The uBorda
algorithm computes a \textit{weighted sample mean} and has been shown to return
the correct estimator with high probability on streaming
contexts~\cite{IruLobPer2020arxiv}.
%
Finally, the balance between intensification and exploration is dynamically
adapted by controlling the variance of the probabilistic Mallows model.

@@

\subsection{Distributions over permutations}
Distributions over permutations are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}\Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\Sigma \assign$ sampled $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}\Comment{Generate $\beta$ permutations}
    \State $\sigma\assign  \arg\max_{\pi'\in \Sigma}\arg\min_{ \pi \in S} d(\pi', \pi)   $ \Comment{Select the farthest from $S$}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}. The algorithm also implements the following strategy to avoid sampling the same points more than once, which will be more likely at the explotation phase. The strategy is as follows. Instead of sampling just one permutation from the MM, the algorithm generates $\beta$ permutations, $\Sigma$. Then it selects  for evaluation the farthest to the sample, i.e.,  $\sigma\in \Sigma$ such that $\sigma = \arg\max\arg\min d(\sigma, \pi)$ for every $\pi\in S$.

@@

\subsubsection{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO
and initially samples $\minit=10$ solutions uniformly. The other two parameters
of UMM are $r_1$ and $r_2$, which determine the balance between exploration and
exploitation. In the experimental section we will study their impact. implemented UMM using Python 3.
At each sampling step, we set $\beta=10$, meaning that 10 permutations are generated and the farthest from $S$ is selected for evaluation.

@@

@@

\subsubsection{Datasets}
%
For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}

@@

In the following paragraphs we analyze the result over real-world instances. %
Figure~\ref{fig:loplib} shows the results on the instances taken from LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
The type of instances from the synthetic LOP instances, and, thus, so does the
landscape. However, UMM shows a very similar behavior: there is a (close to)
linear decrease of the fitness of new solutions over each new evaluation. This
suggests that UMM is able to model the fitness landscape. Regarding CEGO, we
observe a similar behavior for the first 50-100 evaluations. At this point,
CEGO reaches the time limit of 3 days, so no observations can be made for
longer budgets.  Overall, the improvement from the first (random) to the last
solutions is 2-3 times larger in UMM than CEGO. %
%\MANUEL{But this is a bit misleading because UMM runs for 4 times more
%  evaluations, perhaps we should also measure up to the evaluation that CEGO
%  reaches so we report both values?}

@@

%\MANUEL{UMM is able to model the LOP landscape
%  at least as well as CEGO as shown by the plots where UMM shows a linear
%  decrease in fitness.}  
%  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
%  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
%  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
%  luego que addemás es más rápido} 
%  \MANUEL{Pero UMM es peor en rec13 y nug12, y
%  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
%  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
%  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
%  entonces la diferencia en velocidad no importaría. } 
The time required to train the surrogate-model of \ekhine{esta establecida la complexity de CEGO???}{UMM increases linearly with the permutation length $n$, while in }
CEGO  it sharply
increases with $n$, requiring 
sometimes more than hour per fitness evaluation as shown in the experimental section. 
Hence, UMM is a computationally
feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
The results presented here show that, despite the intrinsic difficulty of the
expensive black-box combinatorial scenario, there are still significant
advances yet to be made. Thus, we hope that this paper will motivate further
research. % on such problems.
%\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}
A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's-$\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

@@

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
  \paragraph*{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
  \vspace*{-1em}
\paragraph*{Acknowledgements.}
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable} and to Thomas Stützle (Université libre de Bruxelles) for suggesting works
discussing differences between QAPLIB instances.
Thanks to Ahmed Boujaada for the experimental support. 
\end{smaller}
